Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-10 manuscript.tex
Incorporated new model fits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Different rules for binocular combination of luminance flicker in cortical and subcortical pathways},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within (Baker and Wade, 2017) and between (Ernst and Banks, 2002) the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established (Hubel and Wiesel, 1962). Much of our knowledge about binocular combination derives from studies on the contrast response of the `canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) (Purves et al., 2008). However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels (McDougal and Gamlin, 2008) and much less is known about the computations that operate in these subcortical pathways. In addition, our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between \(\sqrt{2}\) and 2 (Baker et al., 2018; Campbell and Green, 1965). This advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) (Legge, 1984; Meese et al., 2006) and neural activity (Baker and Wade, 2017; Moradi and Heeger, 2009) are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision (Ding and Sperling, 2006; Meese et al., 2006) advocate a process of interocular suppression that normalizes the two eyes inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of `ocularity invariance' (Baker et al., 2007): perceived contrast does not change when one eye is opened and closed.

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (McDougal and Gamlin, 2008). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (1981)). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (Angée et al., 2021; Mathôt, 2018; McDougal and Gamlin, 2008; Wang and Munoz, 2015). To account for the consensual response, these brain regions must combine information from the left and right eyes (Doesschate and Alpern, 1967), yet the computation that achieves this is achieved is unclear.

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural (Norcia et al., 2015) and pupil (Spitschan et al., 2014) responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e.~temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox (Delorme and Makeig, 2004). The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean (see Baker, 2021). For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described by Baker (2021). This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic of Victor and Mast (1991) that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software (Kassner et al., 2014), which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis (e.g. Spitschan et al., 2014). The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies (see Barrionuevo et al., 2014; Barrionuevo and Cao, 2016). Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis (e.g. Figueira et al., 2022).

@@

To describe our data, we chose a model of binocular contrast gain control with the same general form as the first stage of the model proposed by Meese et al. (2006). The second gain control stage was omitted (consistent with Baker and Wade, 2017) to simplify the model and reduce the number of free parameters. The response of the left eye's channel is given by:

@@

Despite being derived from the model proposed by Meese et al. (2006), the simplifications applied to this architecture make it very similar to other models (e.g. Ding and Sperling, 2006; Doesschate and Alpern, 1967; Legge, 1984; Schrödinger, 1926). In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression (see Baker et al., 2012; Kingdom and Libenson, 2015). Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli (Baker and Wade, 2017). Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation (Baker et al., 2020).

We implemented the model within a Bayesian framework using the Stan software (Carpenter et al., 2017). This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\), with these values chosen based on previous literature (Baker et al., 2012; Meese et al., 2006). We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Figure \ref{fig:modelfigure}. We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

Figure \ref{fig:pupildata}c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A \(2\times5\) repeated measures \(ANOVA^2_{circ}\) (Baker, 2021) comparing these conditions revealed a significant main effect of target contrast (F(8,580) = 16.79, \(p\) \textless{} 0.001), a significant effect of condition (F(2,580) = 11.04, \(p\) \textless{} 0.001), and a significant interaction (F(8,580) = 56.25, \(p\) \textless{} 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48\%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(8,328) = 5.34, \(p\) \textless{} 0).

@@

Finally, we calculated the ratio of binocular to monocular responses across the three data types from Experiment 1. Figure \ref{fig:BSratios} shows that these ratios are approximately \(\sqrt2\) across the low-to-intermediate contrast range for all three data types. At higher contrasts, we see ratios of 2 or higher for the EEG data, but much weaker ratios near 1 for the pupillometry data. Note that the ratios here are calculated on a per-participant basis and then averaged, rather than being the ratios of the average values shown in Figures \ref{fig:pupildata} and \ref{fig:EEGdata}. A \(3 \times 5\) repeated measures ANOVA on the logarithmic (dB) ratios found a main effect of contrast (F(2.96,121.48) = 6.18, \emph{p} \textless{} 0), no effect of data modality (F(2,82) = 5.55, \emph{p} = 0.01), but a highly significant interaction (F(5.04,206.46) = 6.38, \emph{p} \textless{} 0).

@@

The strong binocular facilitation and weak interocular suppression in the EEG data from Experiment 1 was very different from previous findings on binocular combination using steady-state EEG with grating stimuli (Baker and Wade, 2017). One possible explanation is that the lower temporal frequency used here (2Hz, vs 5 or 7Hz in previous work) might be responsible for this difference. We therefore ran a second experiment to compare monocular and binocular responses at a range of temporal frequencies. Only EEG data were collected for this experiment, as the pupil response is negligible above around 2Hz (Spitschan et al., 2014); note that we originally chose 2Hz because it produces measurable signals for both EEG and pupillometry, yet is unfortunately optimal for neither.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. Anstis and Ho, 1998; Levelt, 1965; Quaia et al., 2018). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \ref{fig:modelfigure}i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict (Baker and Wade, 2017; Meese et al., 2006), and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage{float} \floatplacement{figure}{H}
\newcommand{\beginsupplement}{\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}}
\usepackage{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Different rules for binocular combination of luminance flicker in cortical and subcortical pathways},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The brain must combine information across multiple sensory inputs to derive a coherent percept of the external world. This involves a process of signal combination both within (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}) and between (\protect\hyperlink{ref-Ernst2002}{Ernst and Banks, 2002}) the senses. Binocular vision is a useful test case for signal combination, as the inputs to the two eyes overlap substantially (in species with forward-facing eyes), and the neural locus is well-established (\protect\hyperlink{ref-Hubel1962}{Hubel and Wiesel, 1962}). Much of our knowledge about binocular combination derives from studies on the contrast response of the `canonical' visual pathway, in which signals pass from the eyes to primary visual cortex (V1), via the lateral geniculate nucleus (LGN) (\protect\hyperlink{ref-Purves2008}{Purves et al., 2008}). However, signals are also combined across the eyes in the network of subcortical nuclei that govern pupil diameter in response to absolute light levels (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}) and much less is known about the computations that operate in these subcortical pathways. In addition, our primary purpose here is to investigate the computations governing signal combination in these two anatomically distinct pathways in response to luminance changes.

For pattern vision, binocular presentation confers greater sensitivity to low contrast targets than monocular presentation. This is known as binocular summation, with summation ratios (the relative improvement under binocular presentation) at detection threshold lying between \(\sqrt{2}\) and 2 (\protect\hyperlink{ref-Baker2018}{Baker et al., 2018}; \protect\hyperlink{ref-Campbell1965}{Campbell and Green, 1965}). This advantage is lost at high stimulus intensities, where both psychophysical performance (contrast discrimination thresholds) (\protect\hyperlink{ref-Legge1984}{Legge, 1984}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}) and neural activity (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Moradi2009}{Moradi and Heeger, 2009}) are approximately equal for monocular and binocular presentation. Contemporary models of binocular vision (\protect\hyperlink{ref-Ding2006}{Ding and Sperling, 2006}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}) advocate a process of interocular suppression that normalizes the two eyes inputs at high contrasts and negates the binocular advantage. This is consistent with our everyday experience of `ocularity invariance' (\protect\hyperlink{ref-Baker2007}{Baker et al., 2007}): perceived contrast does not change when one eye is opened and closed.

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (\protect\hyperlink{ref-Wyatt1981}{1981})). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (\protect\hyperlink{ref-Angee2021}{Angée et al., 2021}; \protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}; \protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}; \protect\hyperlink{ref-Wang2015}{Wang and Munoz, 2015}). To account for the consensual response, these brain regions must combine information from the left and right eyes (\protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}), yet the computation that achieves this is achieved is unclear.

To address this shortcoming, we designed an experiment that allowed us to simultaneously record electrophysiological and pupillometric responses to monocular and binocular stimuli. This novel paradigm allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously in response to flickering light, and make quantitative comparisons between them. Periodic flicker entrains both neural (\protect\hyperlink{ref-Norcia2015}{Norcia et al., 2015}) and pupil (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}) responses at the flicker frequency, enabling precise estimation of response amplitudes in the Fourier domain. We followed up our main experiment with additional exploration of the effect of stimulus frequency, and a psychophysical matching experiment measuring perceived flicker intensity (i.e.~temporal contrast). The results are interpreted using a hierarchical Bayesian computational model of binocular vision, and reveal that subcortical pathways implement stronger interocular suppression than the canonical cortical pathway.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox (\protect\hyperlink{ref-Delorme2004}{Delorme and Makeig, 2004}). The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean (see \protect\hyperlink{ref-Baker2021}{Baker, 2021}). For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described by Baker (\protect\hyperlink{ref-Baker2021}{2021}). This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic of Victor and Mast (\protect\hyperlink{ref-Victor1991}{1991}) that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software (\protect\hyperlink{ref-Kassner2014}{Kassner et al., 2014}), which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis (e.g. \protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}). The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies (see \protect\hyperlink{ref-Barrionuevo2014}{Barrionuevo et al., 2014}; \protect\hyperlink{ref-Barrionuevo2016}{Barrionuevo and Cao, 2016}). Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis (e.g. \protect\hyperlink{ref-Figueira2022}{Figueira et al., 2022}).

@@

To describe our data, we chose a model of binocular contrast gain control with the same general form as the first stage of the model proposed by Meese et al. (\protect\hyperlink{ref-Meese2006}{2006}). The second gain control stage was omitted (consistent with \protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}) to simplify the model and reduce the number of free parameters. The response of the left eye's channel is given by:

@@

Despite being derived from the model proposed by Meese et al. (\protect\hyperlink{ref-Meese2006}{2006}), the simplifications applied to this architecture make it very similar to other models (e.g. \protect\hyperlink{ref-Ding2006}{Ding and Sperling, 2006}; \protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}; \protect\hyperlink{ref-Legge1984}{Legge, 1984}; \protect\hyperlink{ref-Schrodinger1926}{Schrödinger, 1926}). In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression (see \protect\hyperlink{ref-Baker2012}{Baker et al., 2012}; \protect\hyperlink{ref-Kingdom2015}{Kingdom and Libenson, 2015}). Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}). Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation (\protect\hyperlink{ref-Baker2020}{Baker et al., 2020}).

We implemented the model within a Bayesian framework using the Stan software (\protect\hyperlink{ref-Carpenter2017}{Carpenter et al., 2017}). This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\), with these values chosen based on previous literature (\protect\hyperlink{ref-Baker2012}{Baker et al., 2012}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}). We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Figure \ref{fig:modelfigure}. We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

Figure \ref{fig:pupildata}c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A \(2\times5\) repeated measures \(ANOVA^2_{circ}\) (\protect\hyperlink{ref-Baker2021}{Baker, 2021}) comparing these conditions revealed a significant main effect of target contrast (F(8,580) = 16.79, \(p\) \textless{} 0.001), a significant effect of condition (F(2,580) = 11.04, \(p\) \textless{} 0.001), and a significant interaction (F(8,580) = 56.25, \(p\) \textless{} 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48\%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(8,232) = 3.03, \(p\) \textless{} 0.003).

@@

Finally, we calculated the ratio of binocular to monocular responses across the three data types from Experiment 1. Figure \ref{fig:BSratios} shows that these ratios are approximately \(\sqrt2\) across the low-to-intermediate contrast range for all three data types. At higher contrasts, we see ratios of 2 or higher for the EEG data, but much weaker ratios near 1 for the pupillometry data. Note that the ratios here are calculated on a per-participant basis and then averaged, rather than being the ratios of the average values shown in Figures \ref{fig:pupildata} and \ref{fig:EEGdata}. A \(3 \times 5\) repeated measures ANOVA on the logarithmic (dB) ratios found a main effect of contrast (F(3.08,89.28) = 4.53, \emph{p} \textless{} 0.002), no effect of data modality (F(2,58) = 0.75, \emph{p} = 0.48), but a highly significant interaction (F(5.54,160.67) = 3.84, \emph{p} \textless{} \ensuremath{3\times 10^{-4}}).

@@

The strong binocular facilitation and weak interocular suppression in the EEG data from Experiment 1 was very different from previous findings on binocular combination using steady-state EEG with grating stimuli (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}). One possible explanation is that the lower temporal frequency used here (2Hz, vs 5 or 7Hz in previous work) might be responsible for this difference. We therefore ran a second experiment to compare monocular and binocular responses at a range of temporal frequencies. Only EEG data were collected for this experiment, as the pupil response is negligible above around 2Hz (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}); note that we originally chose 2Hz because it produces measurable signals for both EEG and pupillometry, yet is unfortunately optimal for neither.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms (e.g. \protect\hyperlink{ref-Anstis1998}{Anstis and Ho, 1998}; \protect\hyperlink{ref-Levelt1965}{Levelt, 1965}; \protect\hyperlink{ref-Quaia2018}{Quaia et al., 2018}). We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

@@

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \ref{fig:modelfigure}i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict (\protect\hyperlink{ref-Baker2017}{Baker and Wade, 2017}; \protect\hyperlink{ref-Meese2006}{Meese et al., 2006}), and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

\hypertarget{references}{%
\section{References}\label{references}}

@@

\hypertarget{appendix-1}{%
\section{Appendix 1}\label{appendix-1}}

\beginsupplement
